The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-30 manuscript/manuscript.Rnw
use shape instead of attachment

@@

Even though the placenta's primary function is conserved among
mammals, it varies widely in terms of the shape of attachment, degree
of intimacy between fetal tissues and maternal tissues, and degree of
interdigitation of maternal and fetal tissues in different species of
mammals
\cite{Mossman.1987:vertebratefetalmembranes} %,Moffett.ea2006:Immunologyofplacentationin}
(Supplemental \cref{tab:placental_classification}). Mammals also
differ drastically in the number and size of their offspring, as well
as the length of gestation; these traits are often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(Supplemental \cref{tab:gestational_periods}). Although the morphology
of the mammalian placenta has been well characterized in a number of
species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human (for \smonkey) or \mouse (for both Spalacines) transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}

@@

We correlated different aspects of placenta morphology, including
shape, intimacy, and interdigitation with the mRNA transcription
profiles of all 1:1 human orthologs with \ac{FPKM} greater than 10 in
one species, and \ac{FPKM} greater than 1 at least two species.
Supplemental \cref{tab:placental_classification} lists the shape,
intimacy, and interdigitiation of each species analyzed.

\subsubsection{Placenta Attachment Shape}

Supplemental \cref{tab:shape_genes} shows the 50 genes which are most
significantly associated with the shape of placenta attachment, as
well as the shape in which the difference is most significant. The
most significant genes which were associated with diffuse placenta
attachment include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
homolog of \ac{UNC93B1} which is involved in the balancing between
Tlr7 and Tlr9 immune response in
mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
\ac{WNT5A}, which is a member of the Wnt ligand family whose
signaling pathway has been implicated in motility in human trophoblast
cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.

@@

For the degree of intimacy and interdigitation, we saw multiple genes
which were associated with different levels of interdigitiation and
intimacy (Supplemental
\cref{tab:intimacy_genes,tab:interdigitation_genes}) using an
unordered factor approach, but when we used an ordered factor
analysis, none of these genes were retained (Supplemental
\cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
suspect that this is due to the fact that the ancestral state of all
eutherians is a labyrinthine, hemochorial
placenta~\cite{Wildman.ea2006:Evolutionofmammalianplacenta}. This
implies that the transcriptome of intermediately invasive or
interdigitated placentas need not be intermediate between the
transcriptomes of less intimate (folded/lamellar) or interdigitated
placentas (epitheliochorial) and most intimate (hemochorial) or
interdigitated placentas (labyrinthine) because these morphologies
have evolved independently. An alternative is that the limited number
of species with placentas of intermediate interdigitation and intimacy
in this study limits the ability to detect genes which whose
expression correlates with degree of intimacy and interdigitation.
Future analyses of additional placentas with intermediate
interdigitation and intimacy should better illuminate the pathways
which are responsible for these placental morphologies.

>>>>

\listoffixmes

@@

Even though the placenta's primary function is conserved among
mammals, it varies widely in terms of shape, degree
of intimacy between fetal tissues and maternal tissues, and degree of
interdigitation of maternal and fetal tissues in different species of
mammals
\cite{Mossman.1987:vertebratefetalmembranes} %,Moffett.ea2006:Immunologyofplacentationin}
(Supplemental \cref{tab:placental_classification}). Mammals also
differ drastically in the number and size of their offspring, as well
as the length of gestation; these traits are often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(Supplemental \cref{tab:gestational_periods}). Although the morphology
of the mammalian placenta has been well characterized in a number of
species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.

@@

\begin{sloppypar}
Reads quality was evaluated using FASTQC v0.11.2
(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
Transcriptomes without a suitable reference genome (\smonkey, \galili,
\carmeli; italic reference in Supplemental
\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
either the \human (\smonkey) or \mouse (\galili and \carmeli) transcriptome (Supplemental
\cref{tab:ref_genome_aligned_species}), and the
the match with the highest bitscore was chosen as the annotation.
Transcripts where the highest scoring match had a bitscore below 40
were left un-annotated.
% 
All transcriptomes were aligned to the reference (or
assembled) genome using STAR v2.4.2a~%
\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
and expression values in \ac{FPKM}
% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
were
quantified using Cufflinks v2.2.1
\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
%
The genome of the closely-related \textit{P. troglodytes} (separation
$\approx 1$ million years
\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
for \bonobo %
.
\end{sloppypar}

@@

We correlated different aspects of placenta morphology, including
shape, intimacy, and interdigitation with the mRNA transcription
profiles of all 1:1 human orthologs with \ac{FPKM} greater than 10 in
one species, and \ac{FPKM} greater than 1 at least two species.
Supplemental \cref{tab:placental_classification} lists the shape,
intimacy, and interdigitation of each species analyzed.

\subsubsection{Placenta Shape}

Supplemental \cref{tab:shape_genes} shows the 50 genes which are most
significantly associated with the shape of the placenta, as
well as the shape in which the difference is most significant. The
most significant genes which were associated with diffuse placenta
shape include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
homolog of \ac{UNC93B1} which is involved in the balancing between
Tlr7 and Tlr9 immune response in
mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
\ac{WNT5A}, which is a member of the Wnt ligand family whose
signaling pathway has been implicated in motility in human trophoblast
cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.

@@

For the degree of intimacy and interdigitation, we saw multiple genes
which were associated with different levels of interdigitation and
intimacy (Supplemental
\cref{tab:intimacy_genes,tab:interdigitation_genes}) using an
unordered factor approach, but when we used an ordered factor
analysis, none of these genes were retained (Supplemental
\cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
suspect that this is due to the fact that the ancestral state of all
eutherians is a labyrinthine, hemochorial
placenta~\cite{Wildman.ea2006:Evolutionofmammalianplacenta}. This
implies that the transcriptome of intermediately invasive or
interdigitated placentas need not be intermediate between the
transcriptomes of less intimate (folded/lamellar) or interdigitated
placentas (epitheliochorial) and most intimate (hemochorial) or
interdigitated placentas (labyrinthine) because these morphologies
have evolved independently. An alternative is that the limited number
of species with placentas of intermediate interdigitation and intimacy
in this study limits the ability to detect genes which whose
expression correlates with degree of intimacy and interdigitation.
Future analyses of additional placentas with intermediate
interdigitation and intimacy should better illuminate the pathways
which are responsible for these placental morphologies.
